---
title: "Spotify music"
author: "Jonathan Sun"
date: "2022-08-26"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spotify Popularity from 2010 - 2019

This dataset was accessed via DataCamp's Server.
While DataCamp's built-in Workspace supports R, I decided to optimize RStudio's ability to create R Markdown files instead of a Jupyter Notebook just for the sake of it.

The main impetus behind deciding to explore this particular dataset from my background as a musician and as the CEO of a hi-fi audio company.
Even though I am a classically trained violinist, one overlapping factor between genres and music has been the ability for it to move, connect and communicate without borders.
Ysaye's Sonatas and, dare I say, Backstreet Boys all have the capability of making the listener feel.

For these reasons, the tone of this study clearly lean towards the informal side; specifically, I'll be documenting my thought processes and approach in a very colloquial manner.

Despite the levity of the subject matter and my tone, we still will treat the analysis itself seriously, and our process very much remains unchanged.

**The main questions I seek to answer from the dataset are as follows:**

1)  Which genres are the most popular from the last decade?\

    H0: No genre is any more popular than the other\
    HA: One genre is more popular than the other

2)  Does bpm correlate with danceability?\

    H0: No, there is no correlation between bpm and danceability.\

    HA: There is a correlation between bpm and danceability.

3)  Does bpm correlate with popularity?\

    H0: No, there is no correlation between bpm and popularity.\

    HA: Yes, there is a correlation between bpm and popularity.

4)  Is there a particular duration of song that is the most popular?\

    H0: No, there is no duration of song that is the most popular.\

    HA: Yes, there is an ideal song duration for popularity

First we'll need to load the data.
We're going to use the readr library for its speed and quickly examine the first 10 columns, and also the overall shape of the DataFrame.

```{r spotify}
library(readr)
library(dplyr)
library(RColorBrewer)
library(wesanderson)
spotify <- read_csv("spotify.csv")
head(spotify, n=10)
cat(dim(spotify))
```

It seems there's only 549 rows across 14 columns, which is less data than I thought but still a decent amount to work with.

Data is considered tidy format, with each row pertaining to one song, and each column pertaining to an aspect of that song (e.g. title, artist, year, etc.).

Let's look at each column in detail.

```{r}
colnames(spotify)
```

There are 14 columns.
They're broken down as follows:

1)  Title - Title of the song\
2)  Artist - Artist of the song\
3)  Top Genre - The genre the song falls under\
4)  Year - The year the song was released\
5)  BPM - Beats per minute, which is a standard measure of tempo (speed) in a song.\
6)  NRGY - Energy, a subjective measure of a song (higher is faster/louder)\
7)  dnce - Spotify's measure of how danceable a song is (higher is more, lower is less)\
8)  dB - the actual loudness of the song (measured in decibels)\
9)  live - liveness, the chance the song was recorded live\
10) val - Valence, how happy a song is (higher is happy, lower is sad)\
11) dur - the duration of the song\
12) cous - the likeliness the song is acoustic\
13) spch - how much the song has spoken words (higher is more spoken words\
14) pop - Spotify's popularity measure of a song (higher values is more popular)

In order to answer the questions we have, we'd like the following columns (n=10): ['title','artist','genre','year','bpm', 'dnce', 'db', \`'val','dur','pop']

Let's slice the DataFrame to be one with just the columns we need.
**It's important to note here that R does not use zero-indexing, so column 1 is 1 and not 0, and so on.**

```{r}
spotify_slice <- spotify[,c(1,2,3,4,5,7,8,10,11,14)]
print(spotify_slice)
```

It seems like the data is mostly clean (at least based on the limited view we've seen so far).
Let's take a look at whether or not duplicates in the title column or if null values exist anywhere.

```{r}
spotify[duplicated(spotify_slice[,1]),]

```

Well well, looks like there's a few duplicate values (13).
Let's remove these by creating a new variable of a dataframe without these.

```{r}
spotify_slice_no_dup <- spotify_slice[!duplicated(spotify_slice[,1]),]
print(spotify_slice_no_dup)
```

We now have 536 rows.
Let's check to see if there are any NaN values.

```{r}
summary(is.na(spotify_slice_no_dup))
```

Running summary on is.na() demonstrates that all columns have all values.
That makes life a little easier.

Let's now try to answer the first question: **1) Which genres are the most popular from the last decade?**

In order to do so, we will sort the dataset based on popularity, and then genre

```{r}
spotify_top_genres <- table(spotify_slice_no_dup['top genre'])
spotify_top_genres_sorted <- sort(spotify_top_genres, decreasing = TRUE)
spotify_top_genres_sorted_abridged <- tail(sort(spotify_top_genres_sorted, decreasing = FALSE),10)
```

Wow, it looks like dance pop is the most popular genre from 2010-2019, with nearly 300 (298) occurrences!
Don't forget the entire cleaned 2010-2019 dataset only has 536 rows!

Let's visualize that in a graph.

```{r}
#This loads the color package so we can add color to the barplot
library(RColorBrewer)
library(wesanderson)
coul <- brewer.pal(5, "Set2") 
#This creates a horizontal bar plot with col colors set to the palette and labels set horizontally
barplot(spotify_top_genres_sorted_abridged, xlab="Genres", col=coul, horiz=TRUE, las = 1)
```

## 2) Does bpm correlate with danceability?

Next, let's look at if bpm correlates with danceability.

```{r}

```

Once, again let's visualize that.
In order to do so, we will need the help of the package ggpubr

```{r}
library('ggpubr')
ggscatter(spotify_slice_no_dup, x='bpm',y='dnce', add ='reg.line',conf.int = FALSE, cor.coef=TRUE, cor.method='pearson', xlab='BPM', ylab='Danceability')
```

Wait a second here!
According to the data, there's a song with bpm that's nearly 2000 and popularity over 1500.
Clearly there's something wrong here!
Let's remove this outlier.

```{r}
#removing the outlier where bpm is nearly 2000
spotify_slice_no_dup_filtered <- spotify_slice_no_dup[spotify_slice_no_dup[,'bpm']<500,]
print(spotify_slice_no_dup_filtered['bpm'])
```

Let's try plotting again.

```{r}
library('ggpubr')
ggscatter(spotify_slice_no_dup_filtered, x='bpm',y='dnce', add ='reg.line',conf.int = TRUE, cor.coef=TRUE, cor.method='pearson', xlab='BPM', ylab='Danceability')
```

That looks more correct!
It seems as if whether or not a song is fast or slow, has little to do with how danceable it is!
That makes sense.
There's slow dances as well as fast dances!

## 3) Does bpm correlate with popularity?

Once again let's use a scatterplot and visualize this graph.

```{r}
library('ggpubr')
ggscatter(spotify_slice_no_dup_filtered, x='bpm',y='pop', add ='reg.line',conf.int = TRUE, cor.coef=TRUE, cor.method='pearson', xlab='BPM', ylab='Popularity')
```

It also seems as if song temp doesn't seem to matter when it comes to popularity, though most songs seem to cluster around 100 to 150 bpm.


## 4)  Is there a particular duration of song that is the most popular?

This time we'll plot 'duration' (measured in seconds) against 'pop'.

```{r}
library('ggpubr')
ggscatter(spotify_slice_no_dup_filtered, x='dur',y='pop', xlab='Song Duration (seconds)', add ='reg.line',conf.int = TRUE, cor.coef=TRUE, cor.method='pearson', ylab='Popularity')
```
It seems the most popular songs are clustered around 3-4 minutes. There's also a very weak correlation between song length and popularity. 

## Conclusions
Based on our examination of the dataset, it seems like dance pop dominated Spotify's billboards from 2010 to 2019. Bear in mind, this doesn't necessarily mean that dance pop is the best genre as it could be that people who listen to Spotify prefer dance pop. For example, a classical music station like KUSC would likely attract and thus have a higher proportion of classical music fans.

Secondly, it seems like tempo (bpm) has little influence on a song's popularity. This fell within expectations as people like to slow dance (ballads) as well as dance to fast music. 

Thirdly, people did not tend to prefer fast music versus slow. It's important to note here that because there were so many songs that fell within the 'dance pop' genre, that the data could be artificially skewed as such. With that said, the bulk of songs in the dataset were around the 100 to 130 bpm.

Lastly, song duration and popularity were very weakly negatively correlated (r=-0.11). This means that songs that tend to go lower slightly trend towards have lower popularity. However, most music fit within 3-4 minutes, likely because of the influence from dance pop.

In all, though things definitely are not set in stone, chances are that if you're going to listen to a top song on Spotify, it'll likely be a 3-4 minute dance pop song at around 100bpm (Allegretto)!
